---
title: "Debugging/Troubleshooting Tips and Tricks"
description: |
  Help for when things don't turn out how you expect.
author:
  - name: Murray Cadzow 
    url: https://github.com/murraycadzow
    affiliation: University of Otago
    affiliation_url:
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: yes
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Time: 50 min

Description: This session will provide you with advice and skills on how to start tackling error messages and where to look for help.


Errors messages are common when coding and are a way the computer communicates it doesn't understand what the instructions it has been given are.



# Where do I start?

> Finding your bug is a process of confirming the many things that you believe are true — until you find one which is not true.
> —Norm Matloff

My general process focuses on the initial source of the error and then starts to work backwards in command history.

1. Do I recognise the message?
2. Re-look at the command I ran looking for
    - typos
    - missing syntax (e.g. brackets, semicolons, etc.)
    - correct naming of things
3. Is the input for my command what I expect it is?
4. Did the previous command run properly?
    - if not jump to 1. for the previous command


Lets look at a few common R errors 

```{r, eval =  TRUE, error = TRUE}
my_var

mean[2]

a <- c(one = 1,two = 2, three = 3)
a$one

a[[20]]

read.csv("myfile.csv")

ggplot()

notapackage::notafunction()


if(NA){
  print("was NA")
}
```

And examples of warnings

```{r, eva =TRUE, error = TRUE, warning = TRUE}
if(c(3 > c(1,2,3))){
  print("less than three")
}

if(c(1,2,3) > 3){
  print("less than one")
}

mean(c("1", "2"))
```


## Asking for help

- effective googling

Reproducible example








