---
title: "Delving Deeper into R"
description: |
  A deeper look into tpoics that get skipped.
author:
  - name: Murray Cadzow 
    url: https://github.com/murraycadzow
    affiliation: University of Otago
    affiliation_url:
  - name: Matt Bixley
    affiliation: NeSI 
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: yes
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




Time: 90 minutes



Description: This workshop will extend your understanding of R and cover how to interact with non-data.frame structured data such as vectors, matrices and lists. We'll also look at creating your own functions and loops.

Learning objectives:

- data structures
    - homogenous structures
        - vectors
        - matrix
    - heterogenous structures
        - lists
        - data.frame/tibble
- create functions
- create loops
- creating/running scripts


## Data structures

Heavy reliance on content in http://adv-r.had.co.nz/Data-structures.html


### Structures



Homogenous: vector, matrix

```{r}
example_vector <- c(1,2,5,6)
example_vector

example_matrix <- matrix(c(1:10), nrow = 2, byrow = TRUE)
example_matrix
```


Heterogenous: list, data.frame (tibble), data.table(?)


```{r}
example_list <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
example_list
```



- `str`
- `class`
- `attributes`
- `typeof`
- `length`
- `dim`

```{r, eval = FALSE}
items <- c("fork" = 5, "table" = 1,"knife" = 6,"spoon" = 4)
names(items)

attributes(items)

```



### Atomics

- numeric
- logical
- character

Coercion
All elements of an atomic vector must be the same type, so when you attempt to combine different types they will be coerced to the most flexible type. Types from least to most flexible are: 
logical -> integer -> double -> character.

### Subsetting

http://adv-r.had.co.nz/Subsetting.html

Conditional

`<` `>` `==` `!=`


index

`which`


`[` versus `[[` versus `$`


```{r}
example_list

example_list[[c(1,2)]]

example_list[c(1,2)]
```



## Functions

```{r, eval = FALSE}
name <- function(variables) {
  
}
```


## Loops


```{r, eval = FALSE}
for (variable in vector) {
  
}
```


Preallocation:

```{r}

```


