---
title: "Delving Deeper into R"
description: |
  A new article created using the Distill format.
author:
  - name: Murray Cadzow 
    url: https://github.com/murraycadzow
    affiliation: University of Otago
    affiliation_url:
  - name: Matt Bixley
    affiliation: NeSI 
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: yes
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




Time: 90 minutes



Description: This workshop will extend your understanding of R and cover how to interact with non-data.frame structured data such as vectors, matrices and lists. We'll also look at creating your own functions and loops.

Learning objectives:

- data structures
    - homogenous structures
        - vectors
        - matrix
    - heterogenous structures
        - lists
        - data.frame/tibble
- create functions
- create loops
- creating/running scripts


## Data structures

Heavy reliance on content in http://adv-r.had.co.nz/Data-structures.html


### Structures

Homogenous: vector, matrix, array
Heterogenous: list, data.frame (tibble), data.table(?)

- `str`
- `class`
- `attr`
- `length`
- `dim`


### Atomics

- numeric
- logical
- character
- 

## Functions

```{r, eval = FALSE}
name <- function(variables) {
  
}
```


## Loops


```{r, eval = FALSE}
for (variable in vector) {
  
}
```
