---
title: "Useful code snippets for everyday tasks"
description: |
  A new article created using the Distill format.
author:
  - name: Murray Cadzow 
    url: 
    affiliation: University of Otago
    affiliation_url: 
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: yes
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Time: 90 min

Description: This session will cover useful code snippets that are designed to improve your overall experience doing computational research. This will include how to customise your BASH and R environments, useful keyboard short cuts, and short pieces of code to do common tasks e.g. read a directory of files into R.


# R

## Rprofile

```{r, eval = FALSE}

# Load helper packages if using interactive session (doesn't alter your environment)
# Don't add 'analysis' packages here
if (interactive()) {
  suppressMessages(require(devtools))
  suppressMessages(require(usethis))
  suppressMessages(require(testthat))
}

# set CRAN
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# warn on partial matches
options(
  warnPartialMatchArgs = TRUE,
  warnPartialMatchDollar = TRUE,
  warnPartialMatchAttr = TRUE
)

# fancy quotes are annoying and lead to
# 'copy + paste' bugs / frustrations
options(useFancyQuotes = FALSE)
```



## RMarkdown

In rmd the code chunk `eval` option takes more than just `TRUE` or `FALSE` - if you want to selectively run lines within a rmd code chunk you can exclude them explicitly:

````
```{r, eval = c(-1,-3)}`r ''`
1

3

5 # only this line will evaluate
```
````

```{r, eval = c(-1,-3), echo = TRUE}
1

3

5 # only this line will evaluate
```



## Workflow advice

- here

```{r, eval = FALSE}
library(here)

here()
```


- janitor

```{r, eval = FALSE}
library(janitor)

clean_names()
```


- usethis

```{r, eval = FALSE}
library(usethis)

use_r()
edit_r_profile()
create_project()
```


- Rprojects


Read in a directory of files

```{r, eval = FALSE}
library(tidyverse)
files <- list.files(pattern = "*.csv", full.names = TRUE)

my_csvs <- map(files, read_csv)

```


quickly find out the number of blank entries in a column:

```{r, eval = FALSE}
table(is.na(df$colname))
```

replace blank cells in a data.frame with NA:

```{r, eval = FALSE}
data[data == ""] = NA
```

Reorder a vector
```{r}
x <- c("b", "c", "a")

# sorts the vector then returns the indices 
order(x)

x[order(x)]



df <- data.frame(x = rep(1:3, each = 2), y = 6:1, z = letters[1:6])

```



format your numbers into a fixed width (turns from numeric to character)

```
# returns character type of number rounded to 3 decimal places
sprintf('%.3f', 0.123456) 
```

format **all** numeric columns to 3 decimal places and make into a table


```{r}
library(tidyverse)
 mtcars %>% mutate(across(where(is.numeric), list(~sprintf('%.3f', .) ) )) %>% kableExtra::kbl()
```



*tidy* way to transpose a dataframe/tibble

```{r}

mtcars %>%
  tibble::rownames_to_column() %>% # may or may not be needed
  tidyr::pivot_longer(-rowname,
                      names_to = "var", 
                      values_to = "value") %>% 
  tidyr::pivot_wider(names_from = "rowname", 
                     values_from = "value")
```

pull out the nth string after a string split
```
purr::map_chr(stringr::str_split(string_vec, "pattern"), n)
```

```{r purrr}
purrr::map_chr(stringr::str_split(c("chr1","chr2","chr3"), "chr"), 2) # you would get back c("1","2","3")
```

```{r, eval = FALSE}
library(kableExtra)
```



# Bash

## Configuration


### .bashrc or .zshrc 

BASH is a common UNIX commandline or terminal, but for MacOS the default is zsh.

To find out which shell you are using enter this command:
```
echo $0
```

Creating your .bashrc or .zsh

Bash profile for login
`.bashrc` is the common file that controls your bash set up and usually can be found at `~/.bashrc`. Some systems (such as MacOS) also have a file `.bash_profile`. If your system uses the `.bash_profile` file, you can make it refer to `.bashrc` by having this as the contents of `.bash_profile`:

```
[[ -r ~/.bashrc ]] && . ~/.bashrc
```

In the `.bashrc` file it is useful to set a customised prompt, set variables that are useful - e.g. PATH to define where bash looks for installed software - and set up some custom commands (aliases) to make common tasks easier.

Custom prompt
Creating your own prompt in bash can be really useful rather than having a straight `$`. http://ezprompt.net provides a nice way of modifying your prompt and providing the code to add to your `.bashrc`.

Things you might want to do:

- add your username
- add the hostname (the name of the computer)
- add the current directory
- add the full path to the current directory
- have colour

### Exported variables
Exporting variables is a useful way for defining environmental settings. Often this is setting a bash variable to tell programs where to look for things. This website has a few examples of bash variables (https://www.thegeekstuff.com/2010/08/bash-shell-builtin-commands/).

It is useful to include the RSTUDIO_PANDOC variables below.

Rmarkdown Pandoc
A useful one on the server, is defining where R is going to look for pandoc for compiling RMarkdown documents.

I have the following in my .bashrc file

```
export RSTUDIO_PANDOC=/usr/lib/rstudio/bin/pandoc
```

But the location is likely different on your computer. In R, use the command `rmarkdown::find_pandoc()` to find out where the RStudio version of pandoc is located.


Setting this in your `.bashrc is important because there might be another instance of pandoc that is available on your PATH and might cause issues if you run R from the commandline. RSTUDIO_PANDOC is the name that R has specified to use if you want to customise which pandoc is used.


### Better bash history

Bash records your history as it goes but if you are operating across multiple windows it doesn’t work the way you would hope for - e.g. it is only recorded from a single given session, even if you work in multiple. PROMPT_COMMAND is a bash variable that is run as part of running commands. This particular one is designed to time and date stamp commands (not run as root) and their working directory into a daily log file. The logs live in `~/.logs/` so this needs to be made for the command to run `mkdir -p ~/.logs`.

```
export PROMPT_COMMAND='if [ "$(id -u)" -ne 0 ]; then echo "$(date "+%Y-%m-%d.%H:%M:%S") $(pwd) $(history 1)" >> ~/.logs/bash-history-$(date "+%Y-%m-%d").log; fi'
```

If I want to search my logs I can use `grep <command> ~/.logs/*` and it will tell me all the times and directories I ran a command, and how I ran it. The history in these log files is made up of all commands you run on the computer, regardless of how many terminal windows you have open.

Aliases

If you have commands that you're always typing out such as `ls -lrth` it can be useful to create an `alias` for the command that is shorter and easier to type out.

e.g.
```
alias ll="ls -lrth"
```

This can be written in your `.bashrc` so that you can use them in new sessions. These can be very useful but remember they are only available on machines you have been able to customise your `.bashrc`.

# Python

Virtual environments
- conda
- conda create
- conda activate
- conda deactivate
