---
title: "Useful code snippets for everyday tasks"
description: |
  A new article created using the Distill format.
author:
  - name: Murray Cadzow 
    url: 
    affiliation: University of Otago
    affiliation_url: 
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Time: 90 min

Description: This session will cover useful code snippets that are designed to improve your overall experience doing computational research. This will include how to customise your BASH and R environments, useful keyboard short cuts, and short pieces of code to do common tasks e.g. read a directory of files into R.


# R

## RMarkdown

In rmd the code chunk `eval` option takes more than just `TRUE` or `FALSE` - if you want to selectively run lines within a rmd code chunk you can exclude them explicitly:

````
```{r, eval = c(-1,-3)}`r ''`
1

3

5 # only this line will evaluate
```
````

```{r, eval = c(-1,-3), echo = TRUE}
1

3

5 # only this line will evaluate
```



## Workflow advice

- here
- janitor
- usethis
- Rprojects



# Bash

## Configuration


### .bashrc or .zshrc 

BASH is a common UNIX commandline or terminal, but for MacOS the default is zsh.

To find out which shell you are using enter this command:
```
echo $0
```

Creating your .bashrc or .zsh

- path
- alias
- prompt
- ssh config
- tmux


# Python

Virtual environments
- conda
    - conda create
    - conda activate
    - conda deactivate
